Lossy compression of DNA sequencing quality data

### 2015-05-14 manuscript/q-compression.tex
Revert "Added double-precision floating-point."
This reverts commit e882a7695b2951ac2fd01f730ac17cdc05be4c80.

As the cost of sequencing continues to decrease, the rate of sequence
data production is increasing, placing greater demands on storing and
transferring these vast amount of data. Most methods of sequencing
data compression focus on compressing nucleotide information without
any loss of information. Quality data, however, have different
properties than nucleotide data, and methods compressing nucleotide
sequences efficiently do not perform as well on quality
values. Furthermore, although lossless representation is necessary for
nucleotide sequences, it is not an essential requirement for quality
values.

Existing methods for compressing quality values mostly focus on
minimizing the loss of information with less emphasis on the effects
on subsequent analyses. In this paper, we evaluate several different
compression methods for quality values that compromise accuracy for
better storage efficiency, and assess the resulting impacts on common
bioinformatic analyses using sequence read data.

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little negative effect on subsequent
analysis results. The three compression strategies in this study were
able to produce results similar to those obtained with the original
uncompressed quality values when performing quality control, genome
assembly, and alignment of short reads to a reference sequence.

@@

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective, some loss of information
is deemed acceptable if the inferences from analyses are relatively
unaffected. Sequence reads generated by instruments such as an
Illumina HiSeq, the focus of this research due to its preponderance in
genomics, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low-quality tails at the 3$^{\prime}$
end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
sequence reads motivates our exploration of three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
our consideration of an exemplar of each class.~\cite{Canovas:2014fr}
and~\cite{janin2013adaptive} evaluated the effects of lossy
compression on identifying variants within a data set. Here we
describe our research investigating lossy compression of sequence read
quality values, specifically those associated with Illumina
instruments, with the objective to provide some perspective on several
strategies rather than to develop robust high-quality software for
use. We assess the effects of quality value information loss resulting
from compression on subsequent genomic analyses, including read
preprocessing (filtering and trimming), genome assembly, and read
mapping.

@@

Quality values can be binned, and the minimum number of bins that
allows for any distinction among quality values is two; i.e., two
categories of quality: ``good'' and ``bad''. We implement 2-bin
encoding by setting a quality value threshold empirically determined
by the distribution of quality values across reads. Base-calls are
marked ``bad'' if their quality value falls below the first quartile
minus 1.5 $\times$ the interquartile range (IQR), which is the
difference between the first and third quartile. 1.5 $\times$ IQR is
the value used by Tukey's box plot~\citep{mcgill1978variations}. The
main benefit of this approach is that it is completely data-dependent,
and no assumptions regarding the distribution of the quality values
need to be made. We adopt similar ``good''/``bad'' quality value assignments as the preprocessing tool Sickle~\citep{sickle}, which are 40 and 10 respectively.

With 2-bin encoding, binary encoding is possible, allowing us to use a
single bit to represent the quality of a base instead of the standard
8 bits used to store quality values in \textsc{ascii}. An additional
benefit of 2-bin encoding is the potential for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78}.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=3.35in]{profiles_128.eps}}
\caption{Quality profiles obtained by $k$-means clustering on the
  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
  set using $k$ = 128, with each row corresponding to a quality
  profile. Dark to light colors represent low to high quality
  values. It is readily visible that the two most distinctive features
  of quality profiles is their drop-off position and average overall
  quality. One can also see sporadic low-position values in a handful
  of profiles, likely capturing intermittent problems in the
  sequencing process affecting thousands of reads at a
  time.}\label{fig:profiles_128}
\end{figure}

@@

\subsection{Lossy compression is acceptable for some subsequent  
analyses}

It is common to evaluate lossy compression methods in terms of
information loss, and some measure of compression effectiveness.
However, from a pragmatic perspective, it is the consequences for
subsequent analyses that are most important. Our results showed that
some of the compression methods with relatively high information lossy
(\textsc{mse}) are practical, and possibly advantageous, for some
applications. Many bioinformatics tools proved to be robust to
moderate information loss. Passing the decompressed quality values
through quality control software shows that most methods filter nearly
as many bases as using original quality values. Assembly programs
performing sequence alignment use percent similarity scores that are
typically robust to standard sequencing errors.

Downstream applications dictate which lossy compression methods are
acceptable. As noted above, 2-bin encoding is able to retain
information about low-quality base-calls that occur within stretches
of high-quality base-calls, whereas the polynomial regression methods
tend to smooth these quality values. In addition to aligning fewer
reads than the 2-bin encoding, the polynomial regression methods may
also be hindering downstream tools that use the alignment results. A
\textsc{snp} detection pipeline will commonly use the quality values
to record if a potential difference from the reference is the result
of genetic variation or sequencing error. Erroneously marking a base
as higher quality may lead to the base being incorrectly labeled as a
\textsc{snp}. 2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly, and 7th-degree polynomial
regression modeling outperforms it in nearly all metrics. Profiling
performs quite well in all use cases, and has the best performance in
terms of bits/base-call versus error.

\subsection{Potential for operations on compressed data}

Perhaps one of the greatest potential benefits of compressing quality
values is the ability to perform quality control and possibly other
operations directly on the compressed representations of the data. For
example, with profile-based compression, the $k$ profiles can be
evaluated for (pre-)processing operations such as filtering and
trimming, and the operations transitively applied to the entire set of
reads, thus saving substantial computation.

@@

\section{Conclusion}

We have examined several simple and general approaches for lossy
compression of read quality values and their effect on subsequent
analyses. Although most previous examinations of lossy compression
focused primarily on information loss and compression effectiveness,
we show that common bioinformatic analyses are robust to substantial
loss of information in quality values. Some methods, such as
profiling, have yield excellent compression and provide excellent
results in subsequent analyses, and in some cases surpass performance
of the original uncompressed data.

\section{Availability}
Implementations of the compression methods written in Python and R,
and a pipeline to reproduce the results are available at
\url{https://github.com/cmhill/q-compression}

>>>>

As the cost of sequencing continues to decrease, the rate of sequence
data production is increasing, placing greater demands on storing and
transferring these vast amount of data. Most methods of sequencing
data compression focus on compressing nucleotide information without
any loss of information. Quality data, however, have different
properties than nucleotide data, and methods compressing nucleotide
sequences efficiently do not perform as well on quality
values. Furthermore, although lossless representation might be
necessary for nucleotide sequences, it is not an essential requirement
for quality values.

Previous studies of quality value compression have mostly focus on
minimizing the loss of information with less emphasis on the effects
on bioinformatic analyses. In this paper, we evaluate several
different compression methods for quality values, and assess the
resulting impacts on common bioinformatic analyses using sequence read
data: quality control, genome assembly, and alignment of short reads
to a reference sequence.

\section{Results:}
Lossy compression of quality information can greatly decrease memory
requirements with little or no negative, and our results demonstrate
that some compression methods can result in transformed quality values
that are quite useful, and in some cases advantageous, compared to
original uncompressed values.

@@

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
bioinformatic analyses. From this perspective, some loss of
information is deemed acceptable if the inferences from analyses are
relatively unaffected. Sequence reads generated by instruments such as
an Illumina HiSeq, the focus of this research due to its preponderance
in genomics, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low-quality tails at the 3$^{\prime}$
end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
sequence reads motivates our exploration of three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
our consideration of an exemplar of each class.~\cite{Canovas:2014fr}
and~\cite{janin2013adaptive} evaluated the effects of lossy
compression on identifying variants within a data set. Here we
describe our research investigating lossy compression of sequence read
quality values, specifically those associated with Illumina
instruments, with the objective to provide some perspective on several
strategies rather than to develop robust high-quality software for
use. We assess the effects of quality value information loss resulting
from compression on \textsc{dna} sequence data analyses, including
read preprocessing (filtering and trimming), genome assembly, and read
mapping.

@@

Quality values can be binned, and the minimum number of bins that
allows for any distinction among quality values is two; i.e., two
categories of quality: ``good'' and ``bad''. We implement 2-bin
encoding by setting a quality value threshold empirically determined
by the distribution of quality values across reads. Base-calls are
marked ``bad'' if their quality value falls below the first quartile
minus 1.5 $\times$ the interquartile range (IQR), which is the
difference between the first and third quartile; 1.5 $\times$ IQR is
the value used by Tukey's box plot~\citep{mcgill1978variations}. The
main benefit of this approach is that it is completely data-dependent,
and no assumptions regarding the distribution of the quality values
need to be made. We adopt the quality value assignments of the read
preprocessing tool Sickle~\citep{sickle}, which correspond to the 40
for ``good'' and 10 for ``bad''.

With 2-bin encoding, binary encoding is possible, allowing us to use a
single bit to represent the quality of a base instead of the standard
8 bits used to store quality values in \textsc{ascii}. An additional
possible benefit of 2-bin encoding is for increased adjacency of
identical values and repeating patterns, properties that may increase
effectiveness of subsequent compression using established
algorithms~\cite[e.g.,][]{HUFFMAN:1952nr,Ziv77auniversal,
  DBLP:journals/tit/ZivL78}, although this potential benefit is not
evaluated in this study.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[width=3.35in]{profiles_128.eps}}
\caption{Quality profiles obtained by $k$-means clustering on the
  fragment library from \textit{Rhodobacter sphaeroides} 2.4.1 data
  set using $k$ = 128, with each row corresponding to a quality
  profile. Dark to light colors represent low to high quality
  values. It is readily visible that the two most distinctive features
  of quality profiles is their drop-off position and average overall
  quality. Occaional low values occur early in some profiles, likely
  relecting intermittent problems in the sequencing process affecting
  many reads at a time.}\label{fig:profiles_128}
\end{figure}

@@

We have examined several simple and general approaches for lossy
compression of read quality values, and their effect on bioinformatic
analyses. Our results demonstrate that some compression methods can
result in transformed quality values that are quite useful, and in
some cases advantageous, compared to original uncompressed values.

Downstream applications dictate the relative performance lossy
compression methods.  Many bioinformatics tools proved to be robust to
moderate information loss in base-call quality. The use of compression
transformed quality values in quality control demonstrate that several
compression methods result in approximately the same total number
base-calls passing through the combined filtering and trimming
processes as when using original quality values, though the absolute
and relative number of reads filtered and base-calls trimmed may
differ. Genome assembly appears robust to standard sequencing errors,
and compression transformed quality values, particularly those
associated with profiling-based compression, result in assemblies
better than those based on original uncompressed quality values by
most measures.

Among the potential benefits of compressing quality values is the
ability to perform quality control and possibly other operations
directly on the compressed representations of the data. For example,
with profile-based compression, each of the $k$ profiles can be
evaluated for (pre-)processing operations such as filtering and
trimming, and the operations transitively applied to the entire set of
reads, thus saving substantial computation.

@@

Quality values for base-calls in \textsc{dna} sequencing are not
sacrosanct; they are representations of estimates for measurement
(detection) error determined by various mechanisms including signal
processing characteristics and algorithms. The true worth of quality
values lies in the utility they provide in bioinformatic analyses. In
some cases, as demonstrated here, this value is enhanced through the
quality values transformed through compression processes. Thus
appropriate compression of quality values can both reduce associated
memory cost and improve analysis results.

\section{Availability}
Implementations of the compression methods written in Python and
R~\citep{R-Core-Team:2014yq}, and a pipeline to reproduce the results
are available at \url{https://github.com/cmhill/q-compression}
